\chap{Acknowledgements}

\epigraph{It would be nice, she thought wistfully, if someone could say 'thank you' occasionally.}{--- \textsc{Terry Pratchett}\small\textnormal{, Wyrd Sisters}}

A tremendous number of people have been involved -- directly or indirectly -- in the making of this thesis.

First of all, my thanks go out to the thousands of patients who enrolled in the \scanb{} and ABiM studies and chose to donate their biological material. Their donations laid the groundwork for this thesis and through the SCAN-B biobank and dozens of ongoing and future projects will help many patients worldwide.

\textit{Lao Saal}, my main supervisor. He offered me to become a PhD student in his group, provided interesting research opportunities, supported me in my endeavours, and pushed me to push my projects further than I would have done on my own. He also did not mind my frequent ventures into the open source bioinformatics wilds.

\textit{Håkan Axelson} and \textit{Mattias Höglund}, my assistant supervisors, whose input was much appreciated.

\textit{Åke Borg}, for initiating \scanb{}.

\textit{Johan Vallon-Christersson} and \textit{Jari Häkkinen}, for answering my many \scanb{}-related questions.

The \textit{\scanb{} central laboratory} without whose hard work I would not have had data to analyze.

\textit{Niklas Loman}, \textit{Emma Niméus}, \textit{Christian Ingvar}, \textit{Håkan Olsson}, and \textit{Dorthe Grabau} for letting me shadow them in the clinic, during surgery, and in pathology. Research -- particularly the computational variety -- can be very abstract and it is easy to forget that behind every datapoint is or was a human being. Those visits served as a reminder to always keep the patient in mind.

Current and former members of the extended Translational Oncogenomics group: \textit{Yilun ``Alan'' Chen}, \textit{Anthony George}, \textit{Malin Dahlgren}, \textit{Robert Rigo}, \textit{Sergii Gladchuk}, \textit{Heena Saini}, \textit{Yoshiko Nanki}, \textit{Christof Winter}, \textit{Man-Hung Eric Tang}, \textit{Eleonor Olsson}, \textit{Barbara Lettiero}, \textit{Stefano Misino}, \textit{Willow Hight-Warburton}, \textit{Michelle Lee}, \textit{Madeline Dixon}, \textit{Marina Villamor}, \textit{Sofia Gruvberger-Saal}, and \textit{Jill Howlin}.

\textit{Daniel Filipazzi} and \textit{Anders Kvist}, for keeping the servers running as long as possible (RIP \textit{casa3}, \textit{primolo\{5--12\}}, and \textit{prime2}).

\textit{Björn Frostner}, \textit{Emelie Jalgén}, and \textit{Susanne André} for administrative support.

All co-authors for their input, ideas, and work.

\textit{Björn Canbäck}, for interesting discussions and teaching opportunities.

\textit{John och Augusta Persson stiftelse}, \textit{Maggie Stephens stiftelse}, the \textit{Royal Physiographical Society in Lund}, and the \textit{Lund University Medical Faculty PhD Student Travel Fund} for funding my conference travels.

The open source software community that makes the whole field of bioinformatics possible. In particular I would like to thank the hundreds of authors whose software contributed to this thesis, and the \textit{Bioconda} and \textit{conda-forge} communities for making this software available in a sane way.

The \textit{FreeBSD}, \textit{Biopython}, \textit{Bioconda}, and \textit{conda-forge} projects that allowed me to engage in productive procrastination.

\textit{Merrilyn} and \textit{David Hooley}, and the extended Hooley family for their support.

\textit{Gabi} and \textit{Markus}, my siblings, who bought a Commodore Plus/4 home computer in 1986 that planted the seed for my interest in computing.

\textit{Hermann} und \textit{Maria}, meine Eltern. Danke für alles.

\textit{Emma}, for keeping me sane and/or caffeinated.

Lastly I would like to thank all Medicon Village staff who maintained and refilled the coffee machines, without whom this thesis would not have been possible.

\vspace{1em}
\begin{flushright}
\textit{\myName \\ Lund, December 2020}
\end{flushright}